Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)

### 2018-02-28 paper.tex
update

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Abstracting abstract machines (AAM) is a systematic methodology 
  for constructing abstract interpreters that derived from concrete
  small-step abstract machines. Recent progress applies the same idea 
  on definitional interpreter and obtains a big-step 
  abstracting definitional interpreter (ADI) written in monadic style.
  Danvy shows the correspondence between concrete abstract machines and interpreters.
  But the relations between small-step abstracting abstract machines and 
  big-step abstracting definitional interpreters is not well studied.

  In this paper, we show their correspondence and how to syntactically transform small-step 
  abstracting abstract machines into big-step abstracting definitional 
  interpreters.
  The transformations include fusing, disentangling, refunctionalizing 
  and monadification (or un-CPS to direct-style). All of them properly handle the 
  collecting semantics and the non-determinism of abstract interpretation. 
  After each transformation, we also obtain an intermediary form of 
  abstract interpreter.
  Remarkably, follow the idea that evaluation contexts are defunctionalized 
  continuations, we reveal how precise call/return match in control-flow 
  analysis can be obtained by refunctionalizing a small-step abstracting 
  abstract machine,
  as well as explain how it been lost by defunctionalizing the higher-order
  continuations into an firts-order data type.

@@

\begin{lstlisting}
sealed trait Expr
case class Binding(x: String, e: Expr)

case class Var(x: String) extends Expr
case class App(e1: Expr, e2: Expr) extends Expr
case class Lam(x: String, body: Expr) extends Expr
case class Let(x: String, e: App, body: Expr) extends Expr
\end{lstlisting}

@@

@@

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Abstracting abstract machines (AAM) is a systematic methodology 
  for constructing abstract interpreters that derived from concrete
  small-step abstract machines. Recent progress applies the same idea 
  on definitional interpreter and obtains a big-step 
  abstracting definitional interpreter (ADI) written in monadic style.
  Danvy shows the correspondence between concrete abstract machines and interpreters.
  Yet, the relations between small-step abstracting abstract machines and 
  big-step abstracting definitional interpreters is not well studied.

  In this paper, we show their correspondence and how to syntactically transform small-step 
  abstracting abstract machines into big-step abstracting definitional 
  interpreters.
  The transformations include fusing, disentangling, refunctionalizing 
  and monadification (or un-CPS to direct-style). All of them properly handle the 
  collecting semantics and the non-determinism of abstract interpretation. 
  After each transformation, we also obtain an intermediary form of 
  abstract interpreter.
  Remarkably, refunctionalization sequentializes the abstract evaluation.
  And follow the idea that evaluation contexts are defunctionalized 
  continuations, we reveal how precise call/return match in control-flow 
  analysis can be obtained by refunctionalizing a small-step abstracting 
  abstract machine,
  as well as explain how it been lost by defunctionalizing the higher-order
  continuations into an firts-order data type.

@@

\begin{lstlisting}
sealed trait Expr
case class Var(x: String) extends Expr
case class App(e1: Expr, e2: Expr) extends Expr
case class Lam(x: String, body: Expr) extends Expr
case class Let(x: String, e: App, body: Expr) extends Expr
\end{lstlisting}

@@

\begin{lstlisting}
case class Config(e: Expr, env: Env)
case class ValStore(ds: Set[Storable], store: BStore, time: Time)
type Cache = Store[Config, ValStore]
type RetType = (Set[ValStore], Cache)
type Continue = Function5[Set[ValStore], BStore, Time, Cache, Cache, RetType]

def tick(e: Expr, oldtime: Time): Time = (e::oldtime).take(k)

def aval(e: Expr, env: Env, bstore: BStore, time: Time,
         inCache: Cache, _outCache: Cache, continue: Continue): RetType = {
  val config = Config(e, env)
  if (_outCache.contains(config)) {
    return continue(_outCache(config), bstore, List(), inCache, _outCache)
  }
  val outCache = _outCache.update(config, inCache.getOrElse(config, Set()))
  val newTime = tick(e, time)
  e match {
    case Let(x, App(f, ae), e) if isAtomic(ae) =>
      val closures = aeval(f, env, bstore).toList
      def nd(closures: List[Storable], accds: Set[Storable], accbstore: BStore, acctime: Time,
              accInCache: Cache, accOutCache: Cache): RetType  = closures match {
        case Nil =>
          // TODO: If we use `newTime` instead of `acctime`, is it m-CFA?
          val baddr = allocBind(x, acctime)
          val newEnv = env + (x -> baddr)
          val newStore = accbstore.update(baddr, accds)
          aval(e, newEnv, newStore, acctime, accInCache, accOutCache,
            (vs: Set[ValStore], accStore: BStore, acctime: Time, accInCache: Cache, accOutCache: Cache) => {
              val newOutCache = accOutCache.update(Config(e, newEnv), vs)
              continue(vs, accStore, acctime, accInCache, newOutCache)
            })
        case Clos(Lam(v, body), env_c)::rest =>
          val baddr = allocBind(v, newTime)
          val newEnv = env_c + (v -> baddr)
          val newStore = accbstore.update(baddr, aeval(ae, env, bstore)) //TODO: which store?
          aval(body, newEnv, newStore, newTime, accInCache, accOutCache,
            (vs: Set[ValStore], accStore: BStore, acctime: Time, accInCache: Cache, accOutCache) => {
              val newOutCache = accOutCache.update(Config(body, newEnv), vs)
              val vsds = vs.map(_.ds).foldLeft(Set[Storable]())(_ ++ _)
              nd(rest, vsds++accds, accStore, acctime, accInCache, newOutCache)
            })
      }
      nd(closures, Set(), bstore, newTime, inCache, outCache)
    case ae if isAtomic(ae) =>
      val ds = aeval(ae, env, bstore)
      continue(Set(ValStore(ds, bstore, newTime)), bstore, newTime, inCache, outCache)
    }
}

def done: Function2[Expr, Env, Continue] = (e: Expr, env: Env) => {
  (vs: Set[ValStore], store: BStore, time: Time, inCache: Cache, outCache: Cache) => {
    val newOutCache = outCache.update(Config(e, env), vs)
    (vs, outCache)
  }
}

def analyze1(e: Expr, cache: Cache): RetType = {
  aval(e, Map(), Store[BAddr, Storable](Map()), List(), cache, cache, done(e, Map()))
}

def analyze(e: Expr): Set[ValStore] = {
  def iter(cache: Cache): RetType = {
    val (vs, newCache) = analyze1(e, cache)
    if (cache == newCache) (vs, newCache)
    else iter(newCache)
  }
  val (vs, cache) = iter(Store[Config, ValStore](Map()))
  val summarized = cache.map.values.map(_.map(_.store).foldLeft(Store[BAddr, Storable](Map()))(_.join(_))).foldLeft(Store[BAddr, Storable](Map()))(_.join(_))
  vs
}
\end{lstlisting}

@@

On equivalence: since refunc interp does not have an explicit concept of $state$,
so 1) we can compare the final states of P4F against with refunc interp,
or 2) we can compare the summarized store of P4F against with summarized store of refunc interp.
